cardinalR: Generating interesting high-dimensional data structures

### 2025-08-24 paper-cardinalR.Rmd
Updated cross reference in sub sections

The `gen_expbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ exponential branches in $2\text{-}D$, with optional noise dimensions to embed the structure in $p\text{-}D$. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches in $p\text{-}D$. The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

@@

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$. The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.

@@

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in $p\text{-}D$. Each branch lies primarily within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

@@

The `gen_orgcurvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in a $p$-dimensional space. Each branch is constructed in a unique or repeated $2\text{-}D$ subspace of the $p$-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining $p - 2$ dimensions contain Gaussian noise.

@@

The `gen_gridcube(n, p)` constructs a regular grid using approximately $n$ points by evenly partitioning each of the $p$ dimensions. Specifically, the number of grid points along each axis is determined by `gen_nproduct(n, p)`, which calculates the nearest balanced integer factors whose product is close to $n$. The full grid is generated via `expand_grid()`, giving Cartesian combinations of these evenly spaced positions.

@@

The `gen_unifcube(n, p)` function generates $n$ points uniformly distributed within a $3\text{-}D$ cube centered at the origin, optionally embedded in $p\text{-}D$ with additional noise dimensions.

@@

The `gen_cubehole(n, p)` function generates a dataset of $n$ points sampled uniformly inside a cube, then removes points that fall within a central spherical hole, resulting in a hollow cube structure in $p\text{-}D$.

@@

The `gen_quadratic(n, p, range)` function generates a dataset of $n$ points forming a quadratic curve in the first two dimensions. This $2\text{-}D$ parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The `gen_cubic(n, p, range)` function generates a dataset of $n$ points forming a cubic curve in the first two dimensions. This function creates a more complex curvilinear structure than a simple parabola.

@@

The `gen_pyrrect(n, p, h, l_vec, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in $p\text{-}D$, with a tip at height zero and base at height $h$. The shape tapers linearly from the base dimensions $(l_x, l_y)$ to a smaller rectangular section with side lengths $(2r_t, 2r_t)$ at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

@@

The `gen_pyrtri(n, p, h, l, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a triangular cross-section. The structure is embedded in a $p\text{-}D$, with the tip located at height $0$ and the base at height $h$. The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

@@

The `gen_pyrstar(n, p, h, rb)` function generates a dataset of $n$ points forming a high-dimensional pyramid-shaped structure with a hexagonal base. The pyramid extends vertically from height $z = 0$ (tip) to height $z = h$ (base), embedded in a $p$-dimensional space. The distribution concentrates more points near the base, and the hexagonal spread tapers toward the tip.

@@

The `gen_pyrholes(n, p)` function generates $n$ points embedded in a $p\text{-}D$ simplex using a chaotic attractor-like midpoint algorithm. The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a "Sierpinski-like pyramid" in high dimensions.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function `gen_scurve(n, p)`, which generates $n$ observations in $p\text{-}D$.

@@

To simulate a variant of the S-curve structure with a removed region ("hole"), we define the function `gen_scurvehole(n, p)`. This function produces $n$ observations in $p\text{-}D$ where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

@@

The function `gen_circle(n, p)` generates a $p$-dimensional dataset of $n$ observations, where the first two dimensions form a unit circle, and the remaining dimensions are structured sinusoidal extensions of the angular parameter with progressively smaller scale.

@@

The `gen_curvycycle(n, p)` function generates a $p$-dimensional dataset of $n$ observations lying on a curved closed loop with controlled high-dimensional sinusoidal deviations. The structure forms a one-dimensional nonlinear cycle embedded in higher-dimensional space.

@@

The `gen_unifsphere(n, p, r)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on the surface of a $3\text{-}D$ sphere of radius $r$, with additional Gaussian noise dimensions added when $p > 3$.

@@

The `gen_gridedsphere(n, p)` function generates a $p$-dimensional dataset of approximately $n$ points evenly distributed on the surface of a $(p!-!1)$-sphere (the unit sphere embedded in $\mathbb{R}^p$).

@@

The `gen_clusteredspheres(n, k, p, r, loc)` function generates a synthetic dataset of $n_1 + kn_2$ observations in $p$-dimensional space, consisting of one large sphere of radius $r_1$ and $k$ smaller spheres of radius $r_2$, each centered at a different random location.

@@

The `gen_hemisphere(n, p)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on a $4\text{-}D$ hemisphere, optionally extended with Gaussian noise in additional dimensions when $p > 4$.

@@

The `gen_crescent(n, p)` function generates a $p$-dimensional dataset of $n$ observations based on a $2\text{-}D$ crescent-shaped manifold with optional structured high-dimensional noise.

@@

The `gen_curvycylinder(n, p, h)` function generates a $p$-dimensional dataset of $n$ observations structured as a $3\text{-}D$ cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when $p > 4$.

@@

The `gen_sphericalspiral(n, p, spins)` function simulates a dataset of $n$ observations that form a spiraling path on a spherical surface in the first four dimensions. When extended beyond $4\text{-}D$, structured nonlinear noise dimensions are added to simulate more realistic high-dimensional manifolds.

@@

The `gen_helicalspiral(n, p)` function generates a dataset of $n$ observations forming a helical spiral, embedded in the first four dimensions of $p\text{-}D$. Additional dimensions (if $p > 4$) are filled with structured noise to mimic high-dimensional complexity.

@@

The `gen_conicspiral(n, p, spins)` function generates a dataset of $n$ points forming a conical spiral in the first four dimensions of $p\text{-}D$. The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a $3\text{-}D$ conic helix.

@@

The `gen_nonlinear(n, p, hc, non_fac)` function simulates a non-linear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables.

@@

The function `gen_trefoil4d()` generates a $4\text{-}D$ trefoil knot using two angular parameters, $\theta$ and $\phi$. The knot is constructed on the $3$-sphere ($S^3 \subset \mathbb{R}^4$), with a band of thickness around the knot path, controlled by `steps`.

@@

The function `gen_trefoil3d()` generates a $3\text{-}D$ stereographic projection of the $4\text{-}D$ trefoil knot by mapping each point $(X_1, X_2, X_3, X_4) \in \mathbb{R}^4$ to $(X_1', X_2', X_3') \in \mathbb{R}^3$ using the transformation $X_1' = X_1 / (1 - X_4)$, $X_2' = X_2 / (1 - X_4)$, and $X_3' = X_3 / (1 - X_4)$, while excluding points where $X_4 = 1$ to avoid division by zero.

>>>>

The `gen_expbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ exponential branches in $2\text{-}D$, with optional noise dimensions to embed the structure in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) a"} else { "Figure \\@ref(fig:branch-proj) a"})`). These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) b"} else { "Figure \\@ref(fig:branch-proj) b"})`). The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

@@

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) c"} else { "Figure \\@ref(fig:branch-proj) c"})`). The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.

@@

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) d"} else { "Figure \\@ref(fig:branch-proj) d"})`). Each branch lies primarily within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

@@

The `gen_orgcurvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in a $p$-dimensional space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) e"} else { "Figure \\@ref(fig:branch-proj) e"})`). Each branch is constructed in a unique or repeated $2\text{-}D$ subspace of the $p$-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining $p - 2$ dimensions contain Gaussian noise.

@@

The `gen_gridcube(n, p)` constructs a regular grid using approximately $n$ points by evenly partitioning each of the $p$ dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) a"} else { "Figure \\@ref(fig:cube-proj) a"})`). Specifically, the number of grid points along each axis is determined by `gen_nproduct(n, p)`, which calculates the nearest balanced integer factors whose product is close to $n$. The full grid is generated via `expand_grid()`, giving Cartesian combinations of these evenly spaced positions.

@@

The `gen_unifcube(n, p)` function generates $n$ points uniformly distributed within a $3\text{-}D$ cube centered at the origin, optionally embedded in $p\text{-}D$ with additional noise dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) b"} else { "Figure \\@ref(fig:cube-proj) b"})`).

@@

The `gen_cubehole(n, p)` function generates a dataset of $n$ points sampled uniformly inside a cube, then removes points that fall within a central spherical hole, resulting in a hollow cube structure in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:cube) c"} else { "Figure \\@ref(fig:cube-proj) c"})`).

@@

The `gen_quadratic(n, p, range)` function generates a dataset of $n$ points forming a quadratic curve in the first two dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly) a"} else { "Figure \\@ref(fig:poly-proj) a"})`). This $2\text{-}D$ parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The `gen_cubic(n, p, range)` function generates a dataset of $n$ points forming a cubic curve in the first two dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly) b"} else { "Figure \\@ref(fig:poly-proj) b"})`). This function creates a more complex curvilinear structure than a simple parabola.

@@

The `gen_pyrrect(n, p, h, l_vec, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a rectangular base (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) a"} else { "Figure \\@ref(fig:pyr-proj) a"})`). The pyramid is embedded in $p\text{-}D$, with a tip at height zero and base at height $h$. The shape tapers linearly from the base dimensions $(l_x, l_y)$ to a smaller rectangular section with side lengths $(2r_t, 2r_t)$ at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

@@

The `gen_pyrtri(n, p, h, l, rt)` function generates a dataset of $n$ points forming a high-dimensional pyramid-like structure with a triangular cross-section (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) b"} else { "Figure \\@ref(fig:pyr-proj) b"})`). The structure is embedded in a $p\text{-}D$, with the tip located at height $0$ and the base at height $h$. The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

@@

The `gen_pyrstar(n, p, h, rb)` function generates a dataset of $n$ points forming a high-dimensional pyramid-shaped structure with a hexagonal base (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) c"} else { "Figure \\@ref(fig:pyr-proj) c"})`). The pyramid extends vertically from height $z = 0$ (tip) to height $z = h$ (base), embedded in a $p$-dimensional space. The distribution concentrates more points near the base, and the hexagonal spread tapers toward the tip.

@@

The `gen_pyrholes(n, p)` function generates $n$ points embedded in a $p\text{-}D$ simplex using a chaotic attractor-like midpoint algorithm (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) d"} else { "Figure \\@ref(fig:pyr-proj) d"})`). The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a "Sierpinski-like pyramid" in high dimensions.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function `gen_scurve(n, p)`, which generates $n$ observations in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`).

@@

To simulate a variant of the S-curve structure with a removed region ("hole"), we define the function `gen_scurvehole(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) b"} else { "Figure \\@ref(fig:scurve-proj) b"})`). This function produces $n$ observations in $p\text{-}D$ where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

@@

The function `gen_circle(n, p)` generates a $p$-dimensional dataset of $n$ observations, where the first two dimensions form a unit circle, and the remaining dimensions are structured sinusoidal extensions of the angular parameter with progressively smaller scale (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) a"} else { "Figure \\@ref(fig:sphere-proj) a"})`).

@@

The `gen_curvycycle(n, p)` function generates a $p$-dimensional dataset of $n$ observations lying on a curved closed loop with controlled high-dimensional sinusoidal deviations (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) b"} else { "Figure \\@ref(fig:sphere-proj) b"})`). The structure forms a one-dimensional nonlinear cycle embedded in higher-dimensional space.

@@

The `gen_unifsphere(n, p, r)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on the surface of a $3\text{-}D$ sphere of radius $r$, with additional Gaussian noise dimensions added when $p > 3$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) c"} else { "Figure \\@ref(fig:sphere-proj) c"})`).

@@

The `gen_gridedsphere(n, p)` function generates a $p$-dimensional dataset of approximately $n$ points evenly distributed on the surface of a $(p!-!1)$-sphere (the unit sphere embedded in $\mathbb{R}^p$) (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) d"} else { "Figure \\@ref(fig:sphere-proj) d"})`).

@@

The `gen_clusteredspheres(n, k, p, r, loc)` function generates a synthetic dataset of $n_1 + kn_2$ observations in $p$-dimensional space, consisting of one large sphere of radius $r_1$ and $k$ smaller spheres of radius $r_2$, each centered at a different random location (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) e"} else { "Figure \\@ref(fig:sphere-proj) e"})`).

@@

The `gen_hemisphere(n, p)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on a $4\text{-}D$ hemisphere, optionally extended with Gaussian noise in additional dimensions when $p > 4$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) f"} else { "Figure \\@ref(fig:sphere-proj) f"})`).

@@

The `gen_crescent(n, p)` function generates a $p$-dimensional dataset of $n$ observations based on a $2\text{-}D$ crescent-shaped manifold with optional structured high-dimensional noise (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) a"} else { "Figure \\@ref(fig:triginometric-proj) a"})`).

@@

The `gen_curvycylinder(n, p, h)` function generates a $p$-dimensional dataset of $n$ observations structured as a $3\text{-}D$ cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when $p > 4$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) b"} else { "Figure \\@ref(fig:triginometric-proj) b"})`).

@@

The `gen_sphericalspiral(n, p, spins)` function simulates a dataset of $n$ observations that form a spiraling path on a spherical surface in the first four dimensions. When extended beyond $4\text{-}D$, structured nonlinear noise dimensions are added to simulate more realistic high-dimensional manifolds (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) c"} else { "Figure \\@ref(fig:triginometric-proj) c"})`).

@@

The `gen_helicalspiral(n, p)` function generates a dataset of $n$ observations forming a helical spiral, embedded in the first four dimensions of $p\text{-}D$. Additional dimensions (if $p > 4$) are filled with structured noise to mimic high-dimensional complexity (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) d"} else { "Figure \\@ref(fig:triginometric-proj) d"})`).

@@

The `gen_conicspiral(n, p, spins)` function generates a dataset of $n$ points forming a conical spiral in the first four dimensions of $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) e"} else { "Figure \\@ref(fig:triginometric-proj) e"})`). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a $3\text{-}D$ conic helix.

@@

The `gen_nonlinear(n, p, hc, non_fac)` function simulates a non-linear $2\text{-}D$ surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) f"} else { "Figure \\@ref(fig:triginometric-proj) f"})`).

@@

The function `gen_trefoil4d()` generates a $4\text{-}D$ trefoil knot using two angular parameters, $\theta$ and $\phi$. The knot is constructed on the $3$-sphere ($S^3 \subset \mathbb{R}^4$), with a band of thickness around the knot path, controlled by `steps` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) a"} else { "Figure \\@ref(fig:trefoil-proj) a"})`).

@@

The function `gen_trefoil3d()` generates a $3\text{-}D$ stereographic projection of the $4\text{-}D$ trefoil knot by mapping each point $(X_1, X_2, X_3, X_4) \in \mathbb{R}^4$ to $(X_1', X_2', X_3') \in \mathbb{R}^3$ using the transformation $X_1' = X_1 / (1 - X_4)$, $X_2' = X_2 / (1 - X_4)$, and $X_3' = X_3 / (1 - X_4)$, while excluding points where $X_4 = 1$ to avoid division by zero (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) b"} else { "Figure \\@ref(fig:trefoil-proj) b"})`).
